---
# For vignette ---------------------------------------------------------------
title:       'Plotting "hyperSpec" with "ggplot2"'
sutitle:     "Package ggplot2 Based Functionality for hyperSpec Objects"
description: "ggplot2: Package ggplot2 based functionality for hyperSpec"
# Authors --------------------------------------------------------------------
author:
  - name: Claudia Beleites^1,2,3,4,5^, Vilmantas Gegzna
    email: chemometrie@beleites.de
    corresponding : yes
    affiliation   : |
        1. DIA Raman Spectroscopy Group, University of Trieste/Italy (2005--2008)
        2. Spectroscopy $\cdot$ Imaging, IPHT, Jena/Germany (2008--2017)
        3. ÖPV, JKI, Berlin/Germany (2017--2019)
        4. Arbeitskreis Lebensmittelmikrobiologie und Biotechnologie, Hamburg University, Hamburg/Germany (2019 -- 2020)
        5. Chemometric Consulting and Chemometrix GmbH, Wölfersheim/Germany (since 2016)
# Document -------------------------------------------------------------------
date:      "`r Sys.Date()`"
output:
    bookdown::html_vignette2:
      toc: yes
      fig_caption: yes
      css:
        - vignette.css
        - style.css
vignette: >
    % \VignetteIndexEntry{Plotting "hyperSpec" with "ggplot2"}
    % \VignetteKeyword{hySpc.ggplot2}
    % \VignetteKeyword{hyperSpec}
    % \VignetteKeyword{ggplot2}
    % \VignettePackage{hySpc.ggplot2}
    % \VignetteDepends{bookdown, hyperSpec, ggplot2, sessioninfo}
    % \VignetteEngine{knitr::rmarkdown}
    % \VignetteEncoding{UTF-8}
# Citations/References -------------------------------------------------------
link-citations: yes
bibliography: resources/hySpc.ggplot2-pkg.bib
biblio-style: plain
csl: elsevier-with-titles.csl
# Pkgdown --------------------------------------------------------------------
pkgdown:
  as_is: true
  set_null_theme: no
---


```{r cleanup-plotting, include = FALSE}
# Clean up to ensure reproducible workspace ----------------------------------
rm(list = ls(all.names = TRUE))
```

```{r setup, include = FALSE}
# Packages -------------------------------------------------------------------
library(hyperSpec)
library(ggplot2)
library(hySpc.ggplot2)

# Functions ------------------------------------------------------------------
source("vignette-functions.R", encoding = "UTF-8")

# Settings -------------------------------------------------------------------
source("vignette-default-settings.R", encoding = "UTF-8")

# Temporaty options ----------------------------------------------------------
# Change the value of this option in "vignette-default-settings.R"
show_reviewers_notes = getOption("show_reviewers_notes", TRUE)
```

```{r bib, echo=FALSE, paged.print=FALSE}
dir.create("resources", showWarnings = FALSE)

knitr::write_bib(
  c(
    "hyperSpec",
    "ggplot2",
    "hySpc.ggplot2"
  ),
  file = "resources/hySpc.ggplot2-pkg.bib"
)
```



# Suggested Packages and Setup {-}


```{block, type="note", echo=TRUE}
**Reproducing the examples in this vignette.**

All spectra used in this manual are installed automatically with package **hyperSpec**.

```

***

> **Terminology**
> Throughout the documentation of the package, the following terms are used:
>
> - **wavelength** indicates any type of spectral abscissa.
> - **intensity**  indicates any type of spectral ordinate.
> - **extra data** indicates non-spectroscopic data.


```{r packages}
library(hyperSpec)
library(ggplot2)
library(hySpc.ggplot2)

theme_set(theme_bw())

set.seed(2020)
```


# Using **_ggplot2_** with **_hyperSpec_** objects {#plots-ggplot2}


Class `hyperSpec`{.r}[`r cite_pkg("hyperSpec")`] objects do not yet directly support plotting with package **ggplot2**[`r cite_pkg("ggplot2")`].
Nevertheless, package **ggplot2** graphics can easily be obtained by using `qplot*()`{.r} equivalents to `hyperSpec::plotspc()`{.r}, `hyperSpec::plotc()`{.r}, and `hyperSpec::plotmap()`{.r}.


## Function `qplotspc()`{.r} {#qplotspc}

Plot spectra with `qplotspc()`{.r} (Fig. \@ref(fig:ggplotspc)).<!-- `as.long.df()`{.r}" -->

```{r include=FALSE}
CAPTION = "Plot spectra with `as.long.df()`{.r}  "
```

```{r ggplotspc, fig.cap=CAPTION}
qplotspc(flu) + aes(colour = c)
```

```{r}
qplotspc(paracetamol, c(2800 ~ max, min ~ 1800)) +
  scale_x_reverse(breaks = seq(0, 3200, 400))
```

Let's look at the other dataset called `faux_cell`.
```{r}
set.seed(1)
faux_cell <- generate_faux_cell()
faux_cell
```

```{r}
set.seed(25)
qplotspc(sample(faux_cell), spc.nmax = 10) + aes(colour = region)
```

Average spectra of each region.

```{r}
qplotspc(
  aggregate(faux_cell, faux_cell$region, mean),
  mapping = aes(x = .wavelength, y = spc, colour = region)
) +
  facet_grid(region ~ .)
```

Mean $\pm$ standard deviation:
```{r}
qplotspc(
 aggregate(faux_cell, faux_cell$region, mean_pm_sd),
 mapping = aes(x = .wavelength, y = spc, colour = region, group = .rownames)
) +
  facet_grid(region ~ .)

```

## Function `qplotc()`{.r} {#qplotc}

This function plots spectroscopic concenration, depth, time-series, etc. profiles.

By default, plots spectrum only at the first wavelength.
```{r}
qplotc(flu)
```

It is better to indicate the wavelength of interest explicitly.
```{r}
qplotc(flu[ , , 410])
```


```{r}
qplotc(flu[ , , 410]) + 
  geom_smooth(method = "lm", formula = y ~ x)
```

## Function `ggplotmap()`{.r} {#ggplotmap}

Fig. \@ref(fig:ggplotmap) shows a map created with package **ggplot2**.

```{r include=FALSE}
CAPTION = "False-colour map with package **ggplot2**  "
```

```{r ggplotmap, fig.cap=CAPTION}
set.seed(7)
faux_cell <- generate_faux_cell()
```

```{r}
qplotmap(faux_cell[ , , 1200])
```

```{r}
qplotmap(faux_cell[ , , 1200]) +
    scale_fill_gradientn("Intensity", colours = matlab.palette())
```

If you have package **viridis** installed, you may try:
```r
qplotmap(faux_cell[ , , 1200]) +
    viridis::scale_fill_viridis("Intensity")
```

The two special columns `.wavelength`{.r} and `.rownames`{.r} contain the wavelength axis and allow to distinguish the spectra.



## Function `qplotmixmap()`{.r} {#qplotmixmap}

This function plots false-colour map with colour mixing for multivariate overlay.

```{r}
set.seed(1)
faux_cell <- generate_faux_cell()
```

Without pre-processing:
```{r}
qplotmixmap(faux_cell[, , c(800, 1200, 1500)],
  purecol = c(matrix = "red", cell = "green", nucleus = "blue")
)
```

With baseline removed:
```{r}
faux_cell_2 <- faux_cell - spc.fit.poly.below(faux_cell)

qplotmixmap(faux_cell_2[, , c(800, 1200, 1500)],
  purecol = c(matrix = "red", cell = "green", nucleus = "blue")
)
```

With some further pre-processing:
```{r}
faux_cell_3 <- faux_cell_2
faux_cell_3 <- sweep(faux_cell_3, 1, apply(faux_cell_3, 1, mean), "/")
faux_cell_3 <- sweep(faux_cell_3, 2, apply(faux_cell_3, 2, quantile, 0.05), "-")

qplotmixmap(faux_cell_3[, , c(800, 1200, 1500)],
  purecol = c(matrix = "red", cell = "green", nucleus = "blue")
)
```


## More General Plotting with **_ggplot2_** {#general-plotting-ggplot2}


For more general plotting, `as.long.df()`{.r} transforms a `hyperSpec`{.r} object into a long-form `data.frame`{.r} that is suitable for `qplot()`{.r}, while `as.t.df()`{.r} produces a `data.frame`{.r} where each spectrum is one column, and an additional first column gives the wavelength.
Long `data.frame`{.r}s can be *very* memory consuming as they are of size $nrow · nwl × (ncol + 2)$ with respect to the dimensions of the `hyperSpec`{.r} object.
Thus, e.g., the `faux_cell`{.r} data set (`r round (object.size(faux_cell) / 1048576, digits = 0)` MB) as `hyperSpec`{.r} object) needs `r round (object.size(as.long.df(faux_cell, rownames = TRUE)) / 1048576, digits = 0)` MB as long-format `data.frame`{.r}.
It is therefore highly recommended to calculate the particular data to be plotted beforehand.


```{r include=FALSE}
CAPTION = "Mean $\\pm$ standard deviation with package **ggplot2**.  "
```

```{r ggplotmeansd, fig.cap=CAPTION, eval=FALSE}
qplotspc(mean(faux_cell)) +
  geom_ribbon(aes(
    ymin = mean + sd,
    ymax = mean - sd,
    y = 0,
    group = NA
  ),
    alpha = 0.25,
    data = as.t.df(mean_sd(faux_cell))
  )
```

Note that `qpotspc()`{.r} specifies aesthetics `y = spc` and `groups = .rownames`, which do not have corresponding columns in the `data.frame` returned by `as.t.df()`{.r}.
These aesthetics must therefore be set manually in the aesthetics definition in `geom_ribbon()`{.r} (or any other `geom_*()`{.r} that uses `as.t.df()`{.r}).
Otherwise, errors occur that object `spc` (and/or `.rownames`) cannot be found.
Cut axes can be implemented by faceting (Fig. \@ref(fig:ggplotspccut)).

```{r include=FALSE}
CAPTION = "Cut spectra with package **ggplot2**.   "
```

```{r ggplotspccut, fig.cap=CAPTION}
qplotspc(paracetamol / 1e4, wl.range = c(min ~ 1800, 2800 ~ max)) +
  scale_x_continuous(breaks = seq(0, 3200, 400))
```


# Session Info {-}

```{r session-info-plotting, paged.print=FALSE}
sessioninfo::session_info("hyperSpec")
```

# References {-}
